

# Before you begin:
Some useful notes before you read the rest of this demo:

* This was demo was created using R Markdown. R Markdown files have the 
file extension .Rmd. This format is useful because it allows users to write
text (like this) interleaved with R code. When I "knit" this document, 
all of the R code I've written in the .Rmd file is run and the output is 
written to a .HTML file, along with this surrounding text. What this means
for you: 

    + You may find it easier to read the HTML file, rather than the .Rmd. It's
    like the difference between reading a webpage and it's source code. 
    
    + If you're interested in the .Rmd because you want to copy and paste some
    code I used, let me explain how this document is formatted. All of the 
    code is written in "chunks": chunks always begin with three "`" marks
    followd by "{r}", and end with three "`" marks. 
    Everything outside of a chunk is just regular text.

* In R, you can add "comments" to your code using the pound (#) sign. You will
see that I have added many comments to my code below to make it easier for you
to understand what is going on. You do not need to be so thorough in your own
code, but it is generally good practice to add comments explaining the 
purpose of the code you are writing. 

* Before you begin doing work in R, I would recommend downloading your data to 
a folder you've created for the first assignment. Set that to be your working
directory. Whenever we want to load data into R, the file paths we provide are
relative to this working directory. Setting your working directory to be the
same as the location as the data means your file path simplifies to the 
filename. 

* At the end of your R session, you will be prompted about whether you want to
save your workspace.
     + *What this means*: R will save your current work environment
(all of the variables you've created) to a special file format called "RData."
The file is hidden from view: it exists as ".RData" in the working directory,
for those who want to look for it. The next time you open up RStudio and set
your working directory to the same folder, it will load this file and all of 
the variables you created will be there.
     + *I do NOT recommend saving your workspace,* as I think that this 
  ".RData" file just takes up space---oftentimes, a LOT of space---on your
  computer. Also, you might initially think saving your data is a good thing,
  but it's not the results that are meant to be saved---it's the code. You
  should be writing code that is *reproducible*. That means that if I run your
  code, starting from reading in the data, I should be able to produce
  the same (or similar, if doing random simulation) results. 

* You will see me use ```c()```, like ```c("black", "red")```. The ```c()```
function just creates a vector with the arguments I give it--the "c" is short
for "concatenate". 

* Quotation marks in R create character strings: as an example from the demo 
below, you can think of ```"iridium"``` as representing the *word* iridium, 
not the variable ```iridium``` (which is actually a long vector of numbers). 

* When turning in homework, add meaningful labels to your plots. 
I'll show you how to do that below.


# Demo using data from experiments on sublimation temperatures

## Calculating descriptive statistics

Today we'll be working with a set of experiments done to measure the heat of
sublimation for the elements iridium and rhodium. Sublimation is the process of
an element passing from solid state to gaseous state directly, i.e. without
becoming liquid first. 

This data is available as part of a supplement to your textbook. This 
corresponds to Chapter 10 exercise 26. 

### Loading the data

The ```scan()``` function is for reading a list of values into R; in other words,
just a single variable, not multicolumn data like an Excel spreadsheet or table. 
```{r}
# read in data
iridium <- scan("iridium.csv")
rhodium <- scan("rhodium.csv")
```

### Calculating summary statistics
R has many built-in functions to make it easy for you to find out basic 
information about your data. For example, we can compute the mean, median, and
standard deviation of our data:
```{r, collapse=TRUE}
median(iridium)

mean(iridium)

sd(iridium)
```

We can also compute quantiles using the ```quantile()``` function. Recall that
the interquartile range is defined to be the 75th percentile - 25th percentile.
```{r, collapse=TRUE}
# Manually compute the IQR
quantile(iridium, 0.75) - quantile(iridium, 0.25)

# Compute using R's IQR function---same result
IQR(iridium)
```

The ```summary()``` function prints a five-number summary of your data, which
does much of what we just did in one step. 
```{r}
summary(iridium)
```
 
 
 
 
## Exploratory plots
### Histograms
*Histograms* show you the distribution of your data. To produce a histogram in R,
apply the ```hist()``` function to a single numeric vector. You may find the
```breaks``` argument helpful: this sets the number of bins in the histogram.
The more bins, the more detailed your histogram will be. For example, this is the
histogram I get when I let R choose the number of bins for me:
```{r}
hist(iridium,
     xlab = "Temp (C)",     # adjust x-axis label
     main = "Sublimation temperature of Iridium")     # adjust main title
```

And this is the histogram produced when I try to force R to use a lot of bins. 
There are only so many distinct data points, so you see it doesn't actually use
40 bins, but still more than what it had before.
```{r}
hist(iridium, breaks = 40,
     xlab = "Temp (C)",     # adjust x-axis label
     main = "Sublimation temperature of Iridium")     # adjust main title
```

### Boxplots
*Boxplots* are useful for comparing the distributions of two groups. 
The center line is for the median of the data. The box spans the 1st quantile
to the 3rd quantile, so the box represents 50% of the data. The whiskers
extend from 1st quantile - 1.5(IQR) to 3rd quantile + 1.5(IQR). Beyond the
whiskers, points are considered outliers, and directly plotted. 

Remember to use the ```names``` and ```ylab``` arguments to add meaningful
labels to your plot. You can also add an x-axis labels with the argument
```xlab``` and a main title with the argument ```main```. See the scatterplot
in the next section for an example. 
```{r}
boxplot(iridium, rhodium, 
        names = c("iridium", "rhodium"),
        ylab = "Sublimation temperature (Celsius)")
```

### Scatterplot
Here, we produce a *scatterplot* of the sublimation temperatures with respect
to the order they appear in the data. Because the data is already sorted
in the order that the experiments were done, this allows us to look at whether
there's an experiment effect on the sublimation results. 
```{r}
ylims <- range(iridium, rhodium) # find the range of the data

plot(iridium, ylim = ylims, 
     type = "l", # specifies the type of plot as a line (l for line) 
     main = "Sublimation temperature of different elements", # main title
     xlab = "Experiment number", # x-axis label
     ylab = "Temp (C)") # y-axis label

# add a red line for the rhodium data
lines(rhodium, col = "red") 

# add a legend
legend("topright", 
       lwd = 1,       # the data are plotted as lines with thickness (lwd) 1,
       col = c("black", "red"),
       legend = c("iridium", "rhodium"),
       bty = "n")     # get rid of the default black border around the legend
                      # just because I think it looks ugly


```

### Quantile-quantile plot
A quantile-quantile plot graphs the quantiles of one distribution against
another. Typically, we are interested in graphing the quantiles of our data
against a theoretical distribution, to understand how closely the data follow
the theoretical distribution. 

This works especially well for distributions in the location-scale family 
(This refers to distributions where, under shifting and scaling (i.e. linear 
transformations), you still have a distribution in that same family. 
For example, if you shift and scale a normal distribution, you still have a 
normal distribution.) because checking whether your data follows, say, a normal
distribution reduces to seeing whether the QQ plot results in a straight line.
Another distribution in this family is the uniform. 

In the plots below, we simulate data from a normal distribution and create
a QQ-plot comparing the data against a normal distribution with mean 0, sd 1
(that's the default in R). Notice that the points fall on a straight line,
just not the y=x line. In particular, they fall on the y = x + 2 line, 
because the quantiles for the simulated data are roughly 2 units bigger 
than the quantiles for the standard normal.

```{r}
## Normal with a higher mean than the theoretical distribution:
# generate normal variables
normal_vars <- rnorm(100, mean = 2, sd = 1)

# plot a histogram:
hist(normal_vars, breaks = 10,
     main = "Histogram of simulated N(2,1) data",
     xlab = "Simulated N(2,1) data")

# create a qq-plot
qqnorm(normal_vars)
abline(a = 0, b = 1) # add y = x ref line
abline(a = 2, b = 1, col = "red") # add the y = x + 2 line in red

# add a legend:
legend("topleft",
       col = c("black", "red"),
       lwd = 1,
       legend = c("y = x", "y = x +2"))
```


Here's another example with data simulated from a beta distribution, and then
scaled to fall outside the interval (0, 1). This code produces generates the
data and then plots a histogram, superimposing the normal density on top. 
Notice that, relative to the normal density it is being compared to, the
simulated data has much greater density in the lower range. It has a long right
tail. 
```{r}
## A distribution with a long right tail:
# generate random variables
set.seed(1234)
beta_vars <- rbeta(100, 2, 15)*5 - 1.5
# add some more data in the tails
beta_vars <- c(beta_vars, 1, 1.5, 1.5, 1.75, 2, 2.5, 3) 

# plot histogram:
hist(beta_vars, breaks = 15, 
     freq = FALSE,     # plot density instead of counts in the bins
     xlim = c(-2, 2),  # set x-axis range
     main = "Histogram of simulated right skew data",
     xlab = "Simulated data")
curve(dnorm, add = TRUE) # add normal density for reference
```

This is reflected in the QQ plot: to look at this pointwise, we can see from
the histogram that the simulated data attains its 50th percentile at a 
much smaller value than the standard normal. Thus, when the 50th percentile of
the data is plotted against the 50th percentile of the normal, it will fall
below the y = x line. This is true for most of the percentiles being plotted:
the percentiles of the data are generally smaller than those for the standard
normal distribution. 
```{r}
# qq-plot:
qqnorm(beta_vars, ylim = c(-2, 2))
abline(a = 0, b = 1) # add y=x ref line
```



# Homework help
Here's how to read in your data for your first homework problem, and format
it so you have two vectors (one with birthweights for smokers, and the other
with birthweights for non-smokers). 
```{r}
# instead of "babiesI.data", make sure you put the correct file path
# for babiesI.data on your computer. 
babies <- read.table("babiesI.data", header=TRUE)

# create a vector called bwt.smokers with birthweights
# of babies born to mothers who smoke
bwt.smokers <- babies$bwt[babies$smoke == 1] 

# create a vector called bwt.nonsmokers with birthweights 
# of babies born to mothers who don't smoke
bwt.nonsmokers <- babies$bwt[babies$smoke == 0]
```

From here on out, try applying ```summary()``` or ```hist()``` to the 
vectors ```bwt.smokers``` and ```bwt.nonsmokers```. You can even try
making a ```boxplot()```, comparing these two groups--refer to my 
iridium/rhodium boxplot from earlier.
